---
title: "7. Let’s get plotting"
author:
  - name: "Mason A. Wirtz"
    url: https://masonwirtz.github.io
    orcid_id: 0000-0002-9408-1993
output:
  distill::distill_article:
    toc: true
---

For (later) reference, you can find [here a list of common geom_ functions](https://ggplot2.tidyverse.org/reference/)

# Exercise 1

Run the following code chunk

```{r}

# libraries
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse,                                     # data manipulation
  palmerpenguins                                 # package containing data frame to plot
)

theme_set(theme_bw(12))                          # set ggplot2 theme

iris = iris                                      # load in iris data frame
Vampires = read_csv("./Data/Vampires.csv")       # load in Vampires data frame

```

## Exercise 1.1

The data frame `iris` contains the two variables `Sepal.Length` and `Sepal.Width`. Both variables are continuous, meaning we can make a scatter plot to show the relationship between the two variables. Using `Sepal.Length` as the x and `Sepal.Width` as the y, plot these two variables using a scatter plot. 

**HINT** You will need the function `geom_point()`. 


<details>
  <summary>Click for Answer</summary>

**SOLUTION**

When plotting with `ggplot2`, we first have to define which data frame contains the two variables. In this case, this is the data frame `iris`. 

```{r}

iris %>%                                     # data 
  ggplot(aes(x = Sepal.Length,               # set x value
             y = Sepal.Width)) +             # set y value
  geom_point()                               # apply geom_point() function

```

</details> 


## Exercise 1.2 

`ggplot2` is extremely flexible, and we can plot just about everything we can imagine (and then some). The data frame `iris` also contains the variable `Species` (which are the 3 species of iris, namely *setosa*, *versicolor* and *virginica*). Let's say we would like to focus on finding a plotting method to classify the species. To do this, we could simply color-code the data points according to which species they belong to.

Building off the code from the previous code chunk, color-code the variables according to which species (variable: `Species`) they belong to.

**HINT** Remember, when we want to assign colors (i.e. AESTHETICS) to a geom, we do this via the mapping function (i.e. the `aes()` function). 


<details>
  <summary>Click for Answer</summary>

**SOLUTION**

Building off the code we have already written, we can assign the colors using the color argument in the `aes()` function. 

```{r}

iris %>%                                        # data
  ggplot(aes(x = Sepal.Length,                  # set x value
             y = Sepal.Width)) +                # set y value
  geom_point(aes(color = Species))              # aesthetics: separate colors for SPECIES

```

Note, play with the color of the plots, see [this document](https://garthtarr.github.io/meatR/ggplot_extensions.html). 

</details> 



## Exercise 1.3

Using the `labs()` function, change the x axis of the plot to read "Sepal Length", the y axis to read "Sepal Width" and the title to read "Iris Scatterplot".

<details>
  <summary>Click for Answer</summary>
  
**SOLUTION**

Using the `labs()` function is very intuitive, all we need to do is add the `labs()` function and specify the x, y and title arguments. REMEMBER when specifying these, the labs need to be in parentheses!

```{r}

iris %>%                                      # data
  ggplot(aes(x = Sepal.Length,                # set x value
             y = Sepal.Width)) +              # set y value
  geom_point(aes(color = Species)) +          # aesthetics: separate colors for SPECIES
  labs(x = "Sepal Length",                    # label x axis
       y = "Sepal Width",                     # label y axis
       title = "Iris Scatterplot")            # title of the plot

```

</details> 



## Exercise 1.4

Using the `ggsave()` function, save this plot as "scatterplotIris" using a RELATIVE PATH to the folder `Figures`. Save the plot as a .png file. 

**HINT** REMEMBER, we have to save our final plot as an object before saving it! To specify the relative path, you will need the `file = ` argument, i.e. ggplot(OBJECT, file = "./Figure/scatterplotIris.png")

<details>
  <summary>Click for Answer</summary>
  
**SOLUTION**

To save a plot, we first need to save this as an object, which the directions specified as "scatterplotIris". So, we first save the plot using the `=` operator. 


```{r}

scatterplotIris =                           # define object
  iris %>%                                  # data                         
  ggplot(aes(x = Sepal.Length,              # set x value
             y = Sepal.Width)) +            # set y value
  geom_point(aes(color = Species))          # aesthetics: separate colors for SPECIES

# save object to FIGURES folder and name object boxplotPenguine.png
ggsave(scatterplotIris, file = "./Figures/boxplotPenguine.png")      

```

</details> 



# Exercise 2

Before beginning with the following exercises, run the following code:

```{r}

penguins = penguins %>% tibble()           # load in data as a atibble

```


## Exercise 2.1

Take a look at the `penguins` data frame (using either `view(penguins)` or simple `penguins`). Let us say we are interested in the extent to which the body mass of the penguins (variable: `body_mass_g`) differs between islands (variable: `islands`). We would like to plot this using a boxplot. 

If we were to plot this in base R, the plot would look like the following:

```{r}

# RUN this code to see what the plot looks like in base R
# you do not need to do anything with the code other than run it
boxplot(penguins$body_mass_g ~ penguins$island)

```

What is the x variable? What is the y variable? What do we need to enter into our ggplot to recreate this base R plot using `ggplot2`?

**HINT** You will need the geom `geom_boxplot`




<details>
  <summary>Click for Answer</summary>
  
**SOLUTION**

As always, we specify our data frame, which in this case is `penguins`. After that, we need to specify our aesthetics arguments, i.e. the x and y arguments. Since we want to divide the boxplot up by islands, this has to be our x argument, since the x axis shows the different groups (i.e. different islands). The y axis on the other hand shows the body mass of the penguins. After we have defined our aesthetics arguments, we can then simply tell ggplot to plot a boxplot by specifying `geom_boxplot()`. Easy, right?!

```{r}

penguins %>%                            # data
  ggplot(aes(x = island,                # x value
             y = body_mass_g)) +        # y value
  geom_boxplot()                        # apply the BOXPLOT geom

```

</details> 


## Exercise 2.2

Alright, so we have now plotted a boxplot. This is great, but boxplots are not necessarily always the best way to visualize data, because the distribution of the groups are not shown. A boxplot only displays the median, and the data's quartiles. This isn't always helpful, so we often want to somehow show the distribution of our data. We can very easily do this in ggplot by simply adding the individual data points.

To display the individual data points, we can add the geom `geom_jitter` (which is similar to `geom_point()`, the only difference being the dots are spread out a little more). Add `geom_jitter()` to the ggplot code from exercise 2.1.


<details>
  <summary>Click for Answer</summary>
  
**SOLUTION**

The great thing about ggplot is that we can stack geoms on top of each other, adding layers to our plot. This makes it very simple to create informative plots. 

To add the individual data points, we can simply add the `geom_point()` function to the plot we have already created. 

```{r}

penguins %>%                               # data
  ggplot(aes(x = island,                   # set x value
             y = body_mass_g)) +           # set y value
  geom_boxplot() +                         # apply BOXPLOT geom
  geom_jitter()                            # apply JITTER geom (scattered data points)

```

</details> 


## Exercise 2.3

Now, let's say that we would like to color code the islands, so that each island on the boxplot has a different color. How could we do this? 

**HINT** Remember, if we want to specify the ASTHETICS (such as color), we need the `aes()` function, and in this case we would like to FILL the boxplots with color. 



<details>
  <summary>Click for Answer</summary>
  
**SOLUTION**

Since we want to define some type of aesthetic, we need the `aes()` function. We can EITHER define this in the `aes()` function in the `ggplot()` function, or in the `geom_boxplot()` function (doesn't make any difference either way). Since we want to fill the boxplots according to the island, we have to specify the `fill` argument in the `aes()` function, as seen below.

```{r}

penguins %>%                              # data
  ggplot(aes(x = island,                  # set x value
             y = body_mass_g)) +          # set y value
  geom_boxplot(aes(fill = island)) +      # apply BOXPLOT geom; sep. colors for ISLAND
  geom_jitter()                           # apply JITTER geom (scattered data points)

```

</details> 


## Exercise 2.4

Using the `ggsave()` function, save this plot as "boxplotPenguine" using a RELATIVE PATH to the folder `Figures`. 

**HINT** REMEMBER, we have to save our final plot as an object before saving it! To specify the relative path, you will need the `file = ` argument. Don't forget the NAME the plot in the file argument!

<details>
  <summary>Click for Answer</summary>
  
**SOLUTION**

To save a plot, we first need to save this as an object, which the directions specified as "boxplotPenguine". So, we first save the plot using the `=` operator. 


```{r}

boxplotPenguine =                           # define object
  penguins %>%                              # data
  ggplot(aes(x = island,                    # set x value
             y = body_mass_g)) +            # set y value
  geom_boxplot(aes(fill = island)) +        # apply BOXPLOT geom; sep. colors for ISLAND
  geom_jitter()                             # apply JITTER geom (scattered data points)

# save boxplotPenguine object in FIGURES folder and named boxplotPenguine.pdf
ggsave(boxplotPenguine, file = "./Figures/boxplotPenguine.pdf")

```

</details> 





# Exercise 3

Especially when we are first having a look at our data, and when we then want to somehow plot our descriptive statistics (e.g. gender, age etc.), many are tempted to use barplots. Barplots, however, are relatively uninformative, and can oftentimes be misleading. In the words of McElreath (2015: 203)

> Barplots suck. [...] The only problem with barplots is that they have bars. The bars carry only a little information—which way to zero, usually—but greatly clutter the presentation and generate optical illusions.

It's therefore better to use other visualization methods such as dotplots, density plots, violin plots, eye or half-eye plots. These types of plots are extremely easy to make in R (even easier than barplots). 

## Exercise 3.1

Let's take our `Vampires` data frame. We would like to visualize the data distribution of age (variable: `ageOfVampire`) in our group using a density plot (`geom_density()`). When plotting a density plot, the `aes()` function only takes an `x = ` argument.  



<details>
  <summary>Click for Answer</summary>

**SOLUTION**

When visualizing a density plot, we as always need to define which data frame our variable(s) are saved in. In this case, this is the `Vampires` data frame. We then use the pipe and begin with the `ggplot()` function. We then need to define our `x = ` argument in the `aes()` function. After, we simply apply the `geom_density()` function.

```{r}

Vampires %>%                            # data
  ggplot(aes(x = ageOfVampire)) +       # set x value
  geom_density()                        # apply DENSITY geom

```

</details> 

## Example 3.2

Alright, now let's say we would like to display the age distribution according to the gender of the vampires. To do this, we can apply the `fill` argument to the `aes()` function, and instructing ggplot to plot the `ageOfVampire` variable, filling the density plots by gender. 

Apply the variable `gender` to the fill argument in the `aes()` function.


<details>
  <summary>Click for Answer</summary>

**SOLUTION**

Now, to plot the density plots according to gender, all we need to do is apply the `fill` argument to the `aes()` function. 

```{r}

Vampires %>%                              # data
  ggplot(aes(x = ageOfVampire,            # set x value                 
             fill = gender)) +            # fill density with colors by GENDER
  geom_density()                          # apply DENSITY geom

```

</details> 

## Exercise 3.3

But WAIT! The plots overlap and we can't see them well. We have several options here---probably the easiest is that we can make the density plots 'lighter', so that we can see both plots and their distributions. We do this by applying the `alpha` argument to the respective geom. 

Apply the `alpha = .4` argument to the `geom_density()` function. After, play with changing the .4 to higher / lower (between 0 and 1) to see what this function does. 



<details>
  <summary>Click for Answer</summary>

**SOLUTION**

To make the density plots more transparent, we can apply the `alpha` argument and adjust the degree of transparency.

```{r}

Vampires %>%                           # data
  ggplot(aes(x = ageOfVampire,         # set x value
             fill = gender)) +         # fill density with colors by GENDER
  geom_density(alpha = .4)             # apply DENSITY geom; transparency 40%

```

</details> 




# Exercise 4: The Ugly Plot

[Here a list of common geom_ functions](https://ggplot2.tidyverse.org/reference/): Use this list to test out different geoms on the Vampires, Iris or Penguine data frames. 

WHO CAN MAKE THE UGLIEST PLOT?! 

Take 15-20 minutes and (either alone or in groups) attempt to make the absolut worst, ugliest, horribly looking and utterly vomit-worthy plot you can manage by stacking geoms on top of each other. 

And don't forget to test out [different ggplot2 THEMES](https://ggplot2.tidyverse.org/reference/ggtheme.html)!!

Save this plot, USING A RELATIVE PATH, in the Figures folder as a .png file under the name uglyPlot_*your initials*.png

Send this to me via E-Mail (mason.wirtz@stud.sbg.ac.at) once you are finished. 






# Advanced: Visualizing GAMs using ggplot2

When working with any type of data, visual analysis should always be your first step. Before deciding on the distribution of a model, it is worth considering whether the nonlinearity of a predictor-outcome relationship is pronounced enough to justify using methods *other* than linear methods. 

If you're ever interested in using generalized additive models (GAMs), it is also entirely possible to visually diagnose whether the data you are dealing with are nonlinear enough to advocate/justify using GAMs. 

Let's say we are interested in visualizing the nonlinear effects of age of the vampires on the number of people a vampire has changed into vampires, with seperate smooths fitted to gender. We can diagnose whether a GAM is perhaps justified by plotting this using the geom_smooth function, specifying the method as "gam" and the formula as y as a function of the smooth term x. This shows us a relatively linear model, so we might think twice before using a complexer model such as a GAM or GAMM. 

```{r}

Vampires %>%                               # data
  ggplot(aes(x = numberChangedToVamp,      # set x value
             y = ageOfVampire)) +          # set y value
  geom_point() +                           # apply POINT geom (data points)
  geom_smooth(aes(color = gender,          # apply SMOOTH geom; sep. color for GENDER
                  fill = gender),          # fill color by GENDER
              method = "gam",              # smoothing method GAM 
              formula = y ~ s(x)) +        # formula for smoothing function; nonlinear x
  scale_color_viridis_d("gender") +        # define color scheme
  scale_fill_viridis_d("gender")           # define color scheme
```

Let's take a look at what would justify using a GAM: 

```{r}

if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  mgcv,                               # GAMs
  gratia                              # plot GAM
)

set.seed(4444)                        # set seed for reproducability  
dat = gamSim(4, n = 400)              # generate nonlinear data with factors

dat %>%                               # data
  ggplot(aes(x = x2,                  # set x value
             y = y)) +                # set y value
  geom_point() +                      # apply POINT geom (data points)
  geom_smooth(aes(color = fac,        # apply SMOOTH geom; color by FAC
                  fill = fac),        # fill by FAC
              method = "gam",         # smoothing method GAM 
              formula = y ~ s(x)) +   # formula for smoothing function; nonlinear x
  scale_color_viridis_d("fac") +      # define color scheme
  scale_fill_viridis_d("fac")         # define color scheme

```

We see very clear nonlinear trends. Let's compare this to output after we have decided to run a GAM. 

```{r}

model =                       # define object
  bam(y ~ s(x2, by = fac),    # y as a nonlinear function of x2; sep. smooths for FAC
  data = dat                  # data argument
)

draw(model) &                 # plot effects, quick and dirty
  theme_bw()                  # change scheme

```

Given there are no extra random effects/factor smooths to be taken into account, this should provide us with the same smooths as shown above. 








