---
title: "3. Baby steps: Basics of coding in RStudio, part 2"
author:
  - name: "Mason A. Wirtz"
    url: https://masonwirtz.github.io
    orcid_id: 0000-0002-9408-1993
output:
  distill::distill_article:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

Vampires = read_csv("./Data/Vampires.csv")

```


# Excercise 1

Alright, so we've gotten to know a few functions. Let's go ahead and review some of the important ones, specifically the ones that we use really often, like the summary statistics. 

Compute the mean, sd, min and max of the `ageOfVampire` variable in the `Vampires` data frame. 

<details>
  <summary>Click for Answer</summary>
  
**SOLUTION** 

```{r}

mean(Vampires$ageOfVampire)
sd(Vampires$ageOfVampire)
min(Vampires$ageOfVampire)
max(Vampires$ageOfVampire)

```

Remember, when we want to reach into a data frame, we need to use the `$` operator. This tells R to reach into a specific data frame and carry out the function on the variable that comes after the `$` operator. 

</details> 



# Exercise 2

Nice job, you're doing fantastic! So, since you were able to complete the last activity, let's kick it up a notch and get to some fun stuff. 

You've had a hard day, because, let's be honest, that's academia. You need some encouragement, but no one is home to tell you how amazing you are. Let's fix this! 

Load in the package called `praise` and then call the `praise()` function. Run this as many times as you need until you feel like the awesomest person on Earth, because you are!! 

<details>
  <summary>Click for Answer</summary>
  
**SOLUTION** 

```{r}

# install.packages("praise")
library(praise)
praise()

```

Remember, if we haven't installed a package yet, we ALWAYS need to run the `install.packages()` function, with the package name in parentheses. After that, we need to LOAD the package, cause otherwise we have 'downloaded' it, but we haven't actually 'opened' it yet.  

</details> 


# Exercise 3

Alright, so now that we know how to load in packages and call some useful functions, what happens if we forget functions, or if we have something we want to do, but don't remember or know a helpful function for this? Well, GOOGLE is our friend! 

In our `Vampires` data frame, we want to know how many male and female vampires there are. There are a few important steps we need to take to do this. 

## Exercise 3.1

FIRST, we NEED to make sure that all of our variables that should be treated as factor vectors are, indeed, factors. If you have read in the Vampires data set, chances are the `gender` variable was saved as a character vector, which we don't want. 

Go on Google and try to find out which function we can use to change a CHARACTER vector in a data frame to a FACTOR vector (googling something like "change character to factor in r" should do the trick). Your GOAL is to change the `gender` variable in the `Vampires` data frame from a CHARACTER vector to a FACTOR vector (you can see if it worked using the `class(Vampires$gender)` function.)




<details>
  <summary>Click for Answer</summary>
  
**SOLUTION** 

So, there are actually a lot of ways to do this, depending on whether you are using tidyverse (we will learn about this in the next section) or base R. I'll show you a Base R example. 

So, in base R, we work with our `$` operator. Since we want to change a vector in a data frame, we will first have to tell R to change a vector in the data frame, which we do by defining a new variable `Vampires$gender`. Since we already have `gender` in the `Vampires` data frame, by defining our variable in this way, it will overwrite the character vector as a factor vector. 

```{r}

Vampires$gender = as.factor(Vampires$gender)

```


</details> 



## Exercise 3.2

Awesome, you're doing so well! So, since we have now changed our `gender` variable to a factor, we can now count the *factor levels*, i.e. how many different levels does the factor have (in our case, I only entered female/male for the sake of simplicity). Use the `table()` function to count the factor levels of the factor `gender` (what do we need to feed into the `table()` function to make it count the factor levels?)

<details>
  <summary>Click for Answer</summary>
  
**SOLUTION** 

You guessed it! All we need to do is enter the classic `Vampires$gender` into the table function to have it count the number of factor levels each. 

```{r}

table(Vampires$gender)

```

</details> 



## Exercise 3.3

Let's use what we just learned to answer the following questions:

(a) How many vampires in the data frame are dead, and how many alive? 

(b) How many vampires were born on each continent? 

(c) How many vampires are married and how many divorced? 




<details>
  <summary>Click for Answer</summary>
  
**SOLUTION** 

(a) How many vampires in the data frame are dead, and how many alive? 

Well, we first have to factor our variable, then use the `table()` function.

```{r}

Vampires$deadOrAlive = as.factor(Vampires$deadOrAlive)
table(Vampires$deadOrAlive)

```


(b) How many vampires were born on each continent? 

```{r}

Vampires$bornIn = as.factor(Vampires$bornIn)
table(Vampires$bornIn)

```


(c) How many vampires are married and how many divorced? 

```{r}

Vampires$maritalStatus = as.factor(Vampires$maritalStatus)
table(Vampires$maritalStatus)

```

Oh my God, why are so many divorced?!


</details> 





# Exercise 4

Alright, we're going to do some really fun statistics, cause why not? 

Try to do the following (feel free to group up for these!)

(a) Which variables in the `Vampires` data frame are NUMERIC? 

(b) Choose two NUMERIC variables and run a correlation using the `cor.test()` function. This is a fantastic chance to use the help environment to find out what you should enter into the `cor.test()` function! *smiley emoji* 

(c) Install and load the package `report`. Go back and save your correlation test as the variable `cor`. Then run `report(cor)` and thank me later. 

(d) INTERMEDIATE: Change the type of the correlation to a Spearman's correlation. 


<details>
  <summary>Click for Answer</summary>
  
**SOLUTION** 

(a) Which variables in the `Vampires` data frame are NUMERIC? 

So, this one is pretty easy, but we need to know the right function to do this! We can easily use the `str()` function, cause this gives us the classes of all variables in a data frame. 

```{r}

str(Vampires)

```


(b) Choose two NUMERIC variables and run a correlation using the `cor.test()` function. This is a fantastic chance to use the help environment to find out what you should enter into the `cor.test()` function! *smiley emoji* 

So, if you know a little about statistics, you know that a correlation is just the strength of the association between two variables. This logically means that we need to enter two variables into the correlation analysis. Let's say we want to know whether income correlates with wellbeing in vampires. REMEMBER, we need to define in which data frames our variables are coming from using the `$` operator!!

```{r}

cor.test(Vampires$wellbeing, Vampires$income)

```

Nice, looks like there is something going on there! (Remember, since the data generated here are generated newly every time I remake the website, the results you get could be entirely different, so don't let this scare you!! If you got a result, then you did it just fine, go you!)

(c) Install and load the package `report`. Go back and save your correlation test as the variable `cor`. Then run `report(cor)` and thank me later. 

This also works for a whole range of models, from simple correlation analyses all the way up to structual equation modeling and even Bayesian mixed-effects modeling. 

```{r}

# install.packages("report")
library(report)

cor = cor.test(Vampires$income, Vampires$wellbeing)
report(cor)

```


(d) INTERMEDIATE: Change the type of the correlation to a Spearman's correlation. 

```{r}

cor.test(Vampires$income, Vampires$wellbeing, method = "spearman")

```



</details> 



