---
title: "Baby steps: Basics of coding in RStudio, part 1"
description: |
  Unit 2
output: distill::distill_article
---


## Presentation

### What is a variable
- A variable (or object, as we refer to them in `R`) is a 'container' that stores data
- A variable (or object/data container) is
    - saved under a name / defined by a **name**
    - contains **values** and/or content 
    - has a **data type** (numeric, character, logical)

### Assigning variables 

We can assign variables in a few different ways 

- Equal operator **=**  

```{r}

variable = 5

```

- Leftward operator **<-** (short cut: option + -)

```{r}

variable <- 5

```

- Rightward operator **->**

```{r}

5 -> variable

```

### Vectors

What is a vector?

A vector...

- is the most common data structure in R
- is a 1-D set of values with the same type of data structure (e.g. numeric, logical, character)

Examples

Vectors can have different data structures

- Numeric vector

```{r, echo = FALSE}

rnorm(n = 5, mean = 12, sd = 1)

```

- Logical vector

```{r, echo = FALSE}

c(1 == 0, 1 == 0, 1 == 1, 1 == 1, 1 == 0)

```

- Character vector

```{r, echo = FALSE}

c("Mason", "Allen", "Wirtz")

```

Creating vectors

We can create vectors and save them as objects (i.e. as variables)

```{r}

v1 = c(1, 2, 3, 4, 5)
# OR
v2 = c(1:5)

```

What we are using here is the concatenate function, it binds individual integers, characters, etc. together to make a vector. 

**CAREFUL** Everything in a single concatenate sequence must have the **SAME** data type

What do you think is going to happen when we call v1 and v2? 

Creating vectors

Let's have a look...

```{r}

v1
v2

```

Basically, what we did was **string together these integers** to make a vector. 

### Data frames

Creating and understanding data frames

Let's go ahead and look at our `Vampires` data frame again 

```{r}

Vampires = read.csv("Vampires.csv", sep = ",") 

```

```{r} 

knitr::kable(head(Vampires), caption = "Example data set: Vampires") # This does not read in the data, just displays the first few rows
DT::datatable(Vampires)

```

**What kind of vectors do we have here in this data frame?**

The Vampires data frame

**idVampire** Integer (Or, more commonly seen saved as a factor, which is stored in R as a vector of integer values with a corresponding set of character values)

**ageOfVampire** Numeric vector

**deadOrAlive** Character vector

**hasFangs** Numeric vector


These vectors, when combined, create a data frame 

